---
title: "Exploratory Analysis - Medical Bill"
author: "Oscar Cassetti"
date: "1/29/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(sqldf)
library(dplyr)
library(ggplot2)
library(pastecs)
```

# Introduction 

This notebook covers the exploratory data analysis for the raw data.

# Input files

I input data files are the following 

1. `bill_amount.csv`   
2. `bill_id.csv`       
3. `clinical_data.csv` 
4. `demographics.csv`


```{r}

billAmount <- read.csv("../data/bill_amount.csv", header=TRUE, stringsAsFactors = FALSE)
bills <- read.csv("../data/bill_id.csv", stringsAsFactors = FALSE)
clinicalData <- read.csv("../data/clinical_data.csv", stringsAsFactors = FALSE)
demographics <- read.csv("../data/demographics.csv", stringsAsFactors = FALSE)

```

First I am going to check if all the data join correctly.

First I am going to check that every bill has an ammount associated with it.

```{r}
source("eda-functions.R")
data.frame(
checks = c(
"Check all bills link an amount",
"Check all bills link to patient clinical data ",
"Check all clinical data links to one or more bill",
"Check all clinical data links to demographics",
"Check all demographic data links to clinical data"
),
results = c(
checkJoinSQL(
"bills b",
"billAmount ba",
"b.bill_id = ba.bill_id" ,
"ba.bill_id is null"
),

checkJoinSQL("bills b", "clinicalData c", "b.patient_id = c.id and b.date_of_admission = c.date_of_admission" , "c.id is null"),

checkJoinSQL(
"clinicalData c",
"bills b",
"b.patient_id = c.id" ,
"b.patient_id is null"
),

checkJoinSQL(
"clinicalData c",
"demographics d",
"d.patient_id = c.id" ,
"d.patient_id is null"
),

checkJoinSQL(
"demographics d",
"clinicalData c",
"d.patient_id = c.id" ,
"c.id is null"
)
)
)

```

Now data quality checks. I start with the billAmount. 
Here I check the distibrution of the bill amount. 
This is also useful to get a good understanding of the data

```{r}

billSummary <- billAmount %>% group_by(bill_id) %>% summarise(amount=sum(amount))
knitr::kable(pastecs::stat.desc(billSummary$amount), digits = 1)

ggplot(billSummary) + geom_density(aes(x=amount)) + ggtitle("Bill amount distribution")

ggplot(billSummary) + geom_boxplot(aes(x="amount", y=amount)) + ggtitle("Bill amount distribution")

```

The amount of the bill has wide range. More dep knowlege would be required to understand if anything above 60K 
are actually outliers.


```{r}
dateFormat <- "%Y-%m-%d"
clinicalData$date_of_admission <- as.Date(clinicalData$date_of_admission, format=dateFormat)
clinicalData$date_of_discharge <- as.Date(clinicalData$date_of_discharge, format=dateFormat)
bills$date_of_admission <- as.Date(bills$date_of_admission, format=dateFormat)
```

Checking on dates

```{r}
ggplot(clinicalData) + geom_density(aes(x=date_of_admission)) + ggtitle("Distribustion of date of adminssion")

ggplot(clinicalData) + geom_density(aes(x=date_of_discharge)) + ggtitle("Distribustion of date of discharge")

ggplot(clinicalData) + geom_density(aes(x=difftime(date_of_discharge, date_of_admission, units = "day"))) +
  xlab("Days")+ 
  ggtitle("Patient length of stay")
```

```{r}
medicalHistoryCols <-  colnames(clinicalData)[grep("medical_history_", colnames(clinicalData))]
preopMedicationCols <- colnames(clinicalData)[grep("preop_medication_", colnames(clinicalData))]
sympthomCols <- colnames(clinicalData)[grep("symptom", colnames(clinicalData))]

medicalColVals <- reshape2::melt(clinicalData[, c("id", medicalHistoryCols)], na.rm = FALSE, id="id", vars=medicalHistoryCols)

medicalColVals$variable <- gsub("medical_history", "med_hist", medicalColVals$variable)

ggplot(medicalColVals) + geom_histogram(aes(x=variable, fill=factor(value)), stat = "count") + ggtitle("Data quality medical history")


preopMedicationVals <- reshape2::melt(clinicalData[, c("id", preopMedicationCols)], na.rm = FALSE, id="id", vars=preopMedicationCols)
preopMedicationVals$variable <- gsub("preop_medication", "pom", preopMedicationVals$variable)

ggplot(preopMedicationVals) + geom_histogram(aes(x=variable, fill=factor(value)), stat = "count") + ggtitle("Data quality medical preop Medications") 

sympthomVals <- reshape2::melt(clinicalData[, c("id", sympthomCols)], na.rm = FALSE, id="id", vars=sympthomCols)


ggplot(sympthomVals) + geom_histogram(aes(x=variable, fill=factor(value)), stat = "count") + ggtitle("Data quality medical sympthoms")


medicalColValsCleaned <- reshape2::melt(clinicalDataCleaned[, c("id", medicalHistoryCols)], na.rm = FALSE, id="id", vars=medicalHistoryCols)

medicalColValsCleaned$variable <- gsub("medical_history", "med_hist", medicalColValsCleaned$variable)

ggplot(medicalColValsCleaned) + geom_histogram(aes(x=variable, fill=factor(value)), stat = "count") + ggtitle("Data quality medical history")

```

Height and weight

```{r}

ggplot(clinicalData) + geom_density(aes(x=height)) + ggtitle("Height distribution")
ggplot(clinicalData) + geom_boxplot(aes(x="height", y=height)) + ggtitle("Height distribution")
ggplot(clinicalData) + geom_density(aes(x=weight)) + ggtitle("Weight distribution")
ggplot(clinicalData) + geom_boxplot(aes(x="weight", y=weight)) + ggtitle("Weight distribution")

ggplot(clinicalData) + geom_density(aes(x=weight/(height/100)^2)) + ggtitle("BMI distribution")

```

```{r}
clinicalDataCleaned <- clinicalData 

table(clinicalData$medical_history_3)

clinicalDataCleaned <- clinicalDataCleaned %>% mutate(
  medical_history_3 = 
    case_when(
      medical_history_3 == "Yes" ~ 1,
      medical_history_3 == "No" ~  0,
      TRUE ~ as.numeric(medical_history_3)
    )
  )

```

I check that the remapping went correctly 

```{r}
table(clinicalData$medical_history_3, clinicalDataCleaned$medical_history_3)
```

Demographics 

```{r}

demographics$date_of_birth <- as.Date(demographics$date_of_birth, format=dateFormat)
ggplot(demographics) + geom_density(aes(x=date_of_birth)) + ggtitle("Distribustion of date of birth")

ageDf <- clinicalData %>% select(id, date_of_admission) %>% left_join(demographics, by=c("id" = "patient_id")) %>% 
  mutate(age=as.numeric(difftime(date_of_admission, date_of_birth, units = "days")/365)) %>% select(age)


ggplot(ageDf) + geom_density(aes(x=age)) + ggtitle("Age at time of admission") 

ggplot(demographics) + geom_histogram(aes(x=gender, fill=gender), stat = "count") + ggtitle("Gender distribution")

ggplot(demographics) + geom_histogram(aes(x=race, fill=race), stat = "count") + ggtitle("Race distribution")

ggplot(demographics) + geom_histogram(aes(x=resident_status, fill=resident_status), stat = "count") + ggtitle("Resident status distribution")
```

From the graph we notice we need to remap some of the fields:

```{r}
demographicsCleaned <- demographics 


demographicsCleaned <- demographicsCleaned %>% mutate(
  gender = factor(
    case_when(
      gender == "f" ~ "Female",
      gender == "m" ~ "Male",
      TRUE ~ as.character(gender)
    )
  ),
  race = factor(
    case_when(
      race == "chinese" ~ "Chinese",
      race == "India" ~ "Indian",
      TRUE ~ as.character(race)
    )
  ),
  resident_status = factor(
    case_when(
      resident_status == "Singapore citizen" ~ "Singaporean",
      TRUE ~ as.character(resident_status)
    )
  )
)

```

We check the distribution post processing 

```{r}
ggplot(demographicsCleaned) + geom_histogram(aes(x=gender, fill=gender), stat = "count") + ggtitle("Gender distribution")

ggplot(demographicsCleaned) + geom_histogram(aes(x=race, fill=race), stat = "count") + ggtitle("Race distribution")

ggplot(demographicsCleaned) + geom_histogram(aes(x=resident_status, fill=resident_status), stat = "count") + ggtitle("Race distribution")


```
```{r}

patientData <- 
sqldf(
sprintf( 
"SELECT 
%s,
%s,
bs.amount
FROM
demographicsCleaned d 
INNER JOIN clinicalDataCleaned c 
  ON c.id = d.patient_id
INNER JOIN (
SELECT 
b.patient_id, 
b.date_of_admission,
SUM(ba.amount) as amount
FROM
 bills b 
INNER JOIN billAmount ba ON ba.bill_id = b.bill_id 
GROUP BY 
b.patient_id, b.date_of_admission
) bs ON bS.patient_id = c.id and bs.date_of_admission = c.date_of_admission
      ", 
paste("d.", colnames(demographicsCleaned), sep="", collapse=","), 
paste("c.", setdiff(colnames(clinicalDataCleaned), colnames(demographicsCleaned)), sep="", collapse=",")))

```

```{r}

patientData <-
  patientData %>% group_by(patient_id) %>% arrange(patient_id, date_of_admission) %>%
  mutate(
  days_since_last_admission = as.numeric(difftime(
  date_of_admission,
  lag(date_of_admission, 1, order_by = date_of_admission),
  units = "days"
  )),
  days_since_last_discharge = as.numeric(difftime(
  date_of_discharge,
  lag(date_of_admission, 1, order_by = date_of_admission),
  units = "days"
  )),
  weight_change = weight - lag(weight, 1, order_by = date_of_admission)
  ) %>% mutate(
  weight_change = if_else(is.na(weight_change), -1, weight_change),
  days_since_last_discharge = if_else(
  is.na(days_since_last_discharge),-1,
  days_since_last_discharge
  ),
  days_since_last_admission = if_else(
  is.na(days_since_last_admission),-1,
  days_since_last_admission
  )
  )

patientDataCounts <-
  patientData %>% group_by(date_of_admission) %>% summarise(t_admission =
  length(patient_id)) %>% arrange(date_of_admission) %>% mutate(
  t_admission_1 = lag(t_admission, 1, order_by = date_of_admission),
  t_admission_2 = lag(t_admission, 2, order_by = date_of_admission),
  t_admission_7 =
  lag(t_admission, 7, order_by = date_of_admission)
  )
  
for(i in 3:5){
  idx <- which(is.na(patientDataCounts[,i]))
  patientDataCounts[idx,i] <- -1  
}

patientData <- patientData %>% group_by(patient_id) %>% mutate(n_admission=1) %>%arrange(patient_id,date_of_admission) %>% mutate(n_admission_csum=cumsum(n_admission))

patientData <- patientData %>% left_join(patientDataCounts, by="date_of_admission") 
```

```{r}
patientData <- as.data.frame(patientData)  
patientData$bmi <- patientData$weight / (patientData$height / 100)^2
patientData$age <- as.numeric(difftime(patientData$date_of_admission, patientData$date_of_birth, units = "days")) /365

patientData$lofs <- as.numeric(difftime(patientData$date_of_discharge, patientData$date_of_admission, units = "days"))

patientData$date_of_admissionYear <- as.numeric(strftime(patientData$date_of_admission, "%Y"))
patientData$date_of_admissionMonth <- as.numeric(strftime(patientData$date_of_admission, "%m"))
patientData$date_of_birthYear <- as.numeric(strftime(patientData$date_of_birth, "%Y"))

for (i in c(medicalHistoryCols, sympthomCols, preopMedicationCols)){
patientData[, i] <- factor(patientData[,i])
}


saveRDS(patientData, "../data/patientData.RDS")


```
```{r}
ggplot(patientData) + geom_density(aes(x=amount))  

IQR(patientData$amount)
quantile(patientData$amount)
spending_level <- cut(patientData$amount, c(0, 50000, 70000, 100000), include.lowest = TRUE, labels = 
c( '[0, 27,0000)', '[27,000, 70,000)' , '[70,000, 100,000]' ))
ggplot(patientData) +  geom_density(aes(x=amount, fill=spending_level)) + ggtitle("Density estimates for various bill amounts")
```

```{r}

library(randomForest)
library(caret)

fitControl <- trainControl(
  method = "repeatedcv",
  number = 10,
  repeats = 5,
  classProbs = TRUE,
  #selectionFunction = "oneSE"
  selectionFunction = "tolerance",
  verboseIter = TRUE,
  summaryFunction = defaultSummary
  )

fitRF <-  train(
  amount~.,
  data=patientData[complete.cases(patientData),  -grep("id$", colnames(patientData))],
  trControl = fitControl,
  metric = "RMSE",
  #nTrain = 0.5,
  nodesize = 2,
  #importance=TRUE,
  ntree = 500,
  do.trace = FALSE,
  verbose = FALSE

  )
saveRDS(fitRF, "fitRf.RDS")
summary(fitRF)
varImpPlot(fitRF$finalModel)


tree <- rpart::rpart(amount~., data=training[, -grep("id$", colnames(patientData))], 
                     control = rpart::rpart.control(minsplit = 20, cp = 0.005) 
                     )

str(training)
rpart::plotcp(tree)

plot(tree)
text(tree)

summary(tree)

x <- predict(tree, training[, -grep("id$", colnames(patientData))]) - training[, 'amount']
hist(x)

fitlm <- lm(amount~.+0, data=training[, -grep("id$", colnames(patientData))])

quantile(patientData$amount)

#Hmisc::rcorr( model.matrix(amount~. ,patientData[, -grep("id$", colnames(patientData))]))

plot(fitlm)
summary(fitlm)
```